### 2025-02-18 paper/paper.tex
Finished editing theory part

One of the primary concerns in high-dimensional regression is multicollinearity—a situation where predictor variables are highly correlated. Multicollinearity leads to inflated standard errors, unstable coefficient estimates, and difficulties in interpreting model parameters. Additionally, high-dimensional models tend to overfit the training data, resulting in poor generalization to unseen observations. Computational complexity further exacerbates these challenges, making it difficult to estimate model parameters efficiently.
To address these issues, researchers have developed dimension reduction techniques that aim to transform high-dimensional predictor spaces into lower-dimensional representations while preserving essential information. Two widely used techniques in this context are Principal Component Regression (PCR) and Partial Least Squares (PLS). These methods construct new predictor variables, referred to as components, that capture the most relevant variation in the dataset.
The use of PCR and PLS is prevalent in various disciplines, including chemometrics, finance, genomics, and environmental science, where datasets often contain a large number of correlated predictors. While both methods offer solutions to the high-dimensional regression problem, they differ in their approach, underlying assumptions, and effectiveness in different scenarios.

@@

    \item Transform the original predictors into principal components:
    \begin{equation}
    Z = X V
    \end{equation}

@@

A key feature of PCR is its ability to reduce variance in estimated coefficients by using only first \( k \) principal components and eliminating components associated with small eigenvalues:
   \begin{equation}
    Z_k = X V_k
    \end{equation}
    \begin{equation}
\hat{\gamma_k} = (Z_k^T Z_k)^{-1} Z_k^T Y
\end{equation}
\begin{equation}
\hat{\gamma_k}  = \Lambda_k^{-1} V_k^T X^T Y
\end{equation}
    where \( \Lambda_k \) contains only the top \( k \) eigenvalues of \( X^T X \), while \( V_k \) consists of the first \( k \) principal components that capture the most variance in \( X \). The selection of \( k \) is often done using cross-validation or based on the cumulative explained variance criterion (Figure 1).

@@

\subsubsection{Conclusion}
Principal Component Regression provides a robust approach for handling multicollinearity and high-dimensional data. By selecting principal components that capture the most variance, PCR ensures stable coefficient estimates with reduced variance. However, its unsupervised nature may lead to suboptimal predictive performance, as it does not consider the response variable in component selection. In the next section, we introduce Partial Least Squares (PLS), which addresses this limitation by incorporating response information into the component selection process.

@@

\subsubsection{Overview}
Partial Least Squares (PLS) is a regression technique specifically designed to handle high-dimensional datasets where multicollinearity and overfitting pose challenges to traditional regression methods. Developed by Herman Wold in the 1960s, PLS integrates elements of dimensionality reduction with regression, making it an effective tool for cases where the number of predictors (\( p \)) is large relative to the number of observations (\( n \)), or where predictors are highly correlated.

@@

\subsubsection{Mathematical Formulation}
In the standard multiple linear regression framework, the relationship between the response \( Y \) and predictor matrix \( X \) is given by:
\begin{equation}
Y = X \beta + \epsilon
\end{equation}
where:
\begin{itemize}
    \item \( Y \) is the \( n \times 1 \) response vector,
    \item \( X \) is the \( n \times p \) predictor matrix,
    \item \( \beta \) is the \( p \times 1 \) coefficient vector to be estimated,
    \item \( \epsilon \) is the \( n \times 1 \) error term, assumed to be normally distributed with mean zero and variance \( \sigma^2 I \).
\end{itemize}

Unlike OLS, which directly estimates \( \beta \) as:
\begin{equation}
\hat{\beta}_{\text{OLS}} = (X^T X)^{-1} X^T Y
\end{equation}
PLS constructs \textbf{latent variables (LVs)} as linear combinations of the original predictors:
\begin{equation}
T = XW, \quad U = YQ
\end{equation}
where:
\begin{itemize}
    \item \( T \) and \( U \) are the latent variables (score matrices) for \( X \) and \( Y \), respectively,
    \item \( W \) and \( Q \) are the weight (loading) matrices that transform \( X \) and \( Y \) into the latent space.
\end{itemize}

PLS selects components by maximizing the covariance between \( XW \) and \( YQ \), ensuring that the extracted components remain highly predictive of the response.

The final regression model is then expressed in terms of these components:
\begin{equation}
Y = T B + \epsilon
\end{equation}
where \( B \) represents the regression coefficients for the latent variables. The estimated regression coefficients for the original predictors can be obtained as:
\begin{equation}
\hat{\beta}_{\text{PLS}} = W (P^T W)^{-1} B
\end{equation}

\subsubsection{Bias-Variance Tradeoff in PLS}
PLS naturally introduces regularization by selecting a reduced set of components. The tradeoff between bias and variance is evident in:
\begin{equation}
\hat{\beta}_{\text{PLS}} = W (P^T W)^{-1} B
\end{equation}
where fewer components lead to lower variance but higher bias. This mechanism prevents overfitting, particularly in high-dimensional settings.

\subsubsection{NIPALS Algorithm}
The \textbf{Nonlinear Iterative Partial Least Squares (NIPALS)} algorithm is an iterative approach used to extract PLS components when the number of predictors is very large.

@@

\subsubsection{SIMPLS Algorithm}
An alternative to NIPALS is the SIMPLS algorithm, which extracts PLS components without iterative deflation, making it computationally more efficient.

@@

SIMPLS is preferred when computational efficiency is a concern, especially for very large datasets.

\subsubsection{Conclusion}
Partial Least Squares Regression offers a robust alternative to standard regression techniques by integrating dimensionality reduction with predictive modeling. By selecting components that maximize covariance with \( Y \), PLS ensures better predictive performance than PCR, which focuses solely on variance preservation in \( X \). Its ability to address multicollinearity and balance the bias-variance tradeoff makes it an ideal method for high-dimensional regression problems.

Compared to PCR, PLS is a supervised approach, ensuring that the extracted components remain relevant for prediction. This makes PLS an effective tool for applications where both dimensionality reduction and accurate predictions are required.

@@

\subsubsection{Elbow Method for Principal Component Regression (PCR)}
In Principal Component Regression (PCR), component selection is based on Principal Component Analysis (PCA), which transforms the original predictor space into a set of orthogonal components that explain the variance in \( X \). Since PCR does not consider the response variable \( Y \) in component selection, it relies on the cumulative explained variance of the predictors.

@@

\subsubsection{Elbow Method for Partial Least Squares (PLS)}
Unlike PCR, which selects components based solely on variance in \( X \), Partial Least Squares (PLS) extracts components that maximize the covariance between \( X \) and \( Y \). Therefore, the Elbow Method in PLS is based on the cumulative explained covariance rather than variance.

@@

\subsubsection{Interpretation and Practical Considerations}
The Elbow Method provides a simple yet effective means of selecting \( k \), but its effectiveness depends on the dataset. If the "elbow" is not clearly visible, alternative methods such as the \textbf{Needle Algorithm} or \textbf{Cross-Validation (CV)} should be used to refine component selection.

In summary:
- For PCR, components are selected based on explained variance, ensuring that the most important directions in \( X \) are retained.
- For PLS, components are selected based on explained covariance, ensuring that the most predictive directions for \( Y \) are retained.

A combination of these methods, along with domain knowledge, should guide the final selection of \( k \) to balance model complexity and predictive performance.

\subsubsection{Needle Algorithm for PCR and PLS}
The Needle Algorithm is an alternative method for determining the optimal number of components in Principal Component Regression (PCR) and Partial Least Squares (PLS). Unlike the Elbow Method, which relies on a visual inspection of a variance or covariance plot, the Needle Algorithm provides a formalized way of detecting the point beyond which additional components contribute negligibly to the model. It is particularly useful when the variance or covariance decay is gradual, making the elbow point ambiguous.

\paragraph{Needle Algorithm for Principal Component Regression (PCR)}
In Principal Component Regression (PCR), the goal is to select the number of components \( k \) that retain most of the variance in \( X \) while avoiding unnecessary complexity. The Needle Algorithm identifies \( k \) by examining the second derivative of the cumulative explained variance function.

Mathematically, the proportion of variance explained by the first \( k \) components is:
\begin{equation}
\text{Explained Variance}(k) = \frac{\sum_{i=1}^{k} \lambda_i}{\sum_{i=1}^{p} \lambda_i}
\end{equation}
where \( \lambda_i \) are the eigenvalues of \( X^T X \).

@@

\paragraph{Needle Algorithm for Partial Least Squares (PLS)}
For Partial Least Squares (PLS), the Needle Algorithm follows the same principle but is applied to the explained covariance between \( X \) and \( Y \) instead of variance in \( X \).

The proportion of covariance explained by the first \( k \) components is given by:
\begin{equation}
\text{Explained Covariance}(k) = \frac{\sum_{i=1}^{k} \text{Cov}(t_i, y)^2}{\sum_{i=1}^{p} \text{Cov}(t_i, y)^2}
\end{equation}
where \( t_i \) are the latent variables extracted by PLS.

@@

\subsubsection{Interpretation and Practical Considerations}
The Needle Algorithm provides a more formal alternative to the Elbow Method, particularly in cases where variance or covariance decays smoothly without a clear "elbow point." However:
\begin{itemize}
    \item \textbf{For PCR}, it helps determine the number of principal components that explain sufficient variance in \( X \).
    \item \textbf{For PLS}, it selects components based on their predictive power by maximizing the covariance between \( X \) and \( Y \).
    \item The threshold \( \epsilon \) should be chosen carefully to balance between model complexity and predictive accuracy.
\end{itemize}
A combination of methods, including the Elbow Method, Needle Algorithm, and Cross-Validation, is recommended to ensure an optimal choice of \( k \).
\subsubsection{Cross-Validation for PCR and PLS}
Cross-Validation (CV) is a data-driven approach to selecting the optimal number of components in Principal Component Regression (PCR) and Partial Least Squares (PLS). Unlike heuristic methods such as the Elbow Method or the Needle Algorithm, which focus on variance or covariance retention, CV directly evaluates the predictive performance of the model by minimizing the generalization error. This ensures that the selected number of components leads to the best tradeoff between bias and variance.

\paragraph{Cross-Validation for Principal Component Regression (PCR)}
In PCR, the key challenge is to determine how many principal components (\( k \)) should be retained to ensure strong predictive performance. Since PCR does not use the response variable \( Y \) when selecting components, cross-validation is crucial in identifying the point where including additional components no longer improves prediction accuracy.

@@

\paragraph{Cross-Validation for Partial Least Squares (PLS)}
Unlike PCR, where components are selected based only on variance in \( X \), PLS constructs components that maximize the covariance between \( X \) and \( Y \). Thus, CV in PLS plays a slightly different role: it helps determine the number of latent variables that maximize predictive accuracy.

@@

Since PLS incorporates \( Y \) in selecting components, it is expected to achieve lower CV-MSE compared to PCR for a given dataset, particularly when \( X \) and \( Y \) are highly correlated.

\subsubsection{Practical Considerations}
While cross-validation is the most rigorous method for selecting \( k \), it has some practical considerations:
\begin{itemize}
    \item \textbf{Computational Cost}: Cross-validation is computationally expensive since it requires training multiple models for different values of \( k \) across multiple folds.
    \item \textbf{Bias-Variance Tradeoff}: Smaller \( k \) values lead to higher bias, while larger \( k \) values lead to higher variance. CV helps find the optimal balance.
    \item \textbf{Choice of \( K \)}: Typically, \( K = 5 \) or \( K = 10 \) is used. A smaller \( K \) reduces computation time, but larger \( K \) provides a more reliable error estimate.
\end{itemize}

\subsubsection{Comparison of PCR and PLS using Cross-Validation}
\begin{itemize}
    \item \textbf{PCR} often requires more components than PLS to achieve the same predictive performance because it does not consider \( Y \) when selecting components.
    \item \textbf{PLS} typically achieves a lower optimal \( k^* \) due to its supervised nature, which aligns components with \( Y \).
    \item Both methods benefit from CV in avoiding overfitting by selecting an appropriate number of components.
\end{itemize}

\subsubsection{Final Recommendation}
Since CV provides a direct measure of predictive performance, it is considered the gold standard for selecting the number of components in both PCR and PLS, therefore CV is applied in our simulations and empirical estimation.

\subsubsection{Comparison of Methods}
\begin{itemize}
    \item The \textbf{Elbow Method} is computationally simple but may be subjective.
    \item The \textbf{Needle Algorithm} provides a more formal criterion but may still require expert judgment.
    \item \textbf{Cross-Validation} is the most robust, as it directly evaluates predictive performance, but it is computationally expensive.
\end{itemize}

\subsection{Final Thoughts on Component Selection}
The choice of \( k \) should be guided by a combination of these methods, balancing the need for variance retention and predictive accuracy. In empirical applications, cross-validation is typically preferred due to its ability to optimize for prediction error directly.
\newpage

@@

\paragraph{Severe Multicollinearity (Low Observations)}
\begin{itemize}
    \item The dataset contains only \( n = 40 \) observations and \( p = 10 \) predictors, creating a scenario where the number of observations is significantly lower than in previous setups.
    \item A base predictor matrix is generated as:
    \[ X_{\text{gen}} \sim \mathcal{N}(0, I) \]
    \item Additional predictors are formed via a linear transformation:
    \[ X_{\text{additional}} = X_{\text{gen}} @ W + \mathcal{N}(0, 0.1) \]
    where \( W \) represents a randomly generated weight matrix, introducing strong collinearity among predictors.
    \item The response variable follows:
    \[ Y = X \beta + \epsilon, \quad \epsilon \sim \mathcal{N}(0, 1.0) \]
    \item Importantly, only \textbf{three} of the 10 predictors actively contribute to generating \( Y \), while the remaining seven are purely collinear noise variables.
\end{itemize}

>>>>

One of the primary concerns in high-dimensional regression is multicollinearity—a situation where predictor variables are highly correlated. Multicollinearity leads to inflated standard errors, unstable coefficient estimates, and difficulties in interpreting model parameters. Additionally, high-dimensional models tend to overfit the training data, resulting in poor generalization to unseen observations. Computational complexity further exacerbates these challenges, making it difficult to estimate model parameters efficiently.

To address these issues, researchers have developed dimension reduction techniques that aim to transform high-dimensional predictor spaces into lower-dimensional representations while preserving essential information. Two widely used techniques in this context are Principal Component Regression (PCR) and Partial Least Squares (PLS). These methods construct new predictor variables, referred to as components, that capture the most relevant variation in the dataset.
The use of PCR and PLS is prevalent in various disciplines, including chemometrics, finance, genomics, and environmental science, where datasets often contain a large number of correlated predictors. While both methods offer solutions to the high-dimensional regression problem, they differ in their approach, underlying assumptions, and effectiveness in different scenarios.

@@

    \item Transform the original predictors into principal components:
    \begin{equation}
    Z = X V
    \end{equation}
    where  \( Z \) represents the uncorrelated principal components of X.

@@

A key feature of PCR is its ability to reduce variance in estimated coefficients by using only first \( k \) principal components and eliminating components associated with small eigenvalues:
   \begin{equation}
    Z_k = X V_k
    \end{equation}

Estimating regression coefficients in the principal component space:    
    \begin{equation}
\hat{\gamma_k} = (Z_k^T Z_k)^{-1} Z_k^T Y
\end{equation}

Using the eigen decomposition (7), we obtain:
\begin{equation}
\hat{\gamma_k}  = \Lambda_k^{-1} V_k^T X^T Y
\end{equation}
    where \( \Lambda_k \) contains only the top \( k \) eigenvalues of \( X^T X \), while \( V_k \) consists of the first \( k \) principal components that capture the most variance in \( X \). The selection of \( k \) is often done using cross-validation or based on the cumulative explained variance criterion (Figure 1).

@@

Principal Component Regression provides a robust approach for handling multicollinearity and high-dimensional data. However, its unsupervised nature may lead to suboptimal predictive performance, as it does not consider the response variable in component selection. In the next section, we introduce Partial Least Squares (PLS), which addresses this limitation by incorporating response information into the component selection process.

@@

Partial Least Squares (PLS) is a regression technique specifically designed to handle high-dimensional datasets where multicollinearity and overfitting pose challenges to traditional regression methods. Developed by Herman Wold in the 1960s, PLS integrates elements of dimensionality reduction with regression, making it an effective tool for cases where the number of predictors (\( p \)) is large relative to the number of observations (\( n \)), or where predictors are highly correlated.

@@

\paragraph {Mathematical Formulation} \ \

PLS constructs \textbf{latent variables (LVs)} as linear combinations of the original predictors:
\begin{equation}
T = XW, \quad U = YQ
\end{equation}
where:
\begin{itemize}
    \item \( T \) and \( U \) are the latent variables (score matrices) for \( X \) and \( Y \), respectively,
    \item \( W \) and \( Q \) are the weight (loading) matrices that transform \( X \) and \( Y \) into the latent space.
\end{itemize}

PLS selects components by maximizing the covariance between \( XW \) and \( YQ \), ensuring that the extracted components remain highly predictive of the response:
\begin{equation}
W = \arg\max_W \operatorname{cov}(XW, Y)
\end{equation}

The final regression model is then expressed in terms of these components:
\begin{equation}
Y = T \delta + \epsilon
\end{equation}
where \( \delta \) represents the regression coefficients for the latent variables.

\paragraph {Bias-Variance Tradeoff in PLS} \ \

Like PCR, PLS reduces variance by using only the first k components. However, PLS retains predictive power better than PCR because it selects components based on their relevance to Y.

Extracting first \( k \) PLS components:
\begin{equation}
T_k = XW_k
\end{equation}
where \(W_k \) contains the first \( k \) PLS weight vectors that capture the most covariance between \( X \) and \( Y \).

Estimating regression coefficients in the latent space:
 \begin{equation}
\hat{\delta_k} = (T_k^T T_k)^{-1} T_k^T Y
\end{equation}

The final regression coefficient in the original space is:
\begin{equation}
\hat{\beta}_{\text{PCR}} = W_k \hat{\delta_k} 
\end{equation}

After substitution we have:
\begin{equation}
\hat{\beta}_{\text{PCR}} = W_k (W_k^T X^T X W_k)^{-1} W_k^T X^T Y
\end{equation}

In the case when all \(p \) components are used, i.e. \(k = p \), we can simplify \(\hat{\beta}_{\text{PCR}} \) and we have:
\begin{equation}
\hat{\beta}_{\text{PLS}} = (X^T X)^{-1} X^T Y = \hat{\beta}_{\text{OLS}} 
\end{equation}
where fewer components lead to lower variance but higher bias. This mechanism prevents overfitting, particularly in high-dimensional settings.

Partial Least Squares Regression offers a robust alternative to standard regression techniques by integrating dimensionality reduction with predictive modeling. Compared to PCR, PLS is a supervised approach, ensuring that the extracted components remain relevant for prediction. This makes PLS an effective tool for applications where both dimensionality reduction and accurate predictions are required.

\paragraph {NIPALS Algorithm} \ \

Several algorithms exist for computing PLS regression. One of the common, The \textbf{Nonlinear Iterative Partial Least Squares (NIPALS)} algorithm is an iterative approach used to extract PLS components when the number of predictors is very large.

@@

\paragraph {SIMPLS Algorithm} \ \

An alternative to NIPALS is the SIMPLS (Statistically Inspired Modification of PLS) algorithm, which extracts PLS components without iterative deflation, making it computationally more efficient.

@@

SIMPLS is preferred when computational efficiency is a concern, especially for very large datasets. Both methods produce the same final PLS regression coefficients, but their computation methods differ.

@@

\paragraph {Elbow Method} \ \

In Principal Component Regression (PCR), component selection is based on Principal Component Analysis (PCA), which transforms the original predictor space into a set of orthogonal components that explain the variance in \( X \). Since PCR does not consider the response variable \( Y \) in component selection, it relies on the cumulative explained variance of the predictors.

@@

Unlike PCR, which selects components based solely on variance in \( X \), Partial Least Squares (PLS) extracts components that maximize the covariance between \( X \) and \( Y \). Therefore, the Elbow Method in PLS is based on the cumulative explained covariance rather than variance.

@@

The Elbow Method provides a simple yet effective means of selecting \( k \), but its effectiveness depends on the dataset. If the "elbow" is not clearly visible, alternative methods such as the \textbf{Needle Algorithm} or \textbf{Cross-Validation (CV)} should be used to refine component selection.

\paragraph {Needle Algorithm} \ \
The Needle Algorithm is an alternative method for determining the optimal number of components, which provides a formalized way of detecting the point beyond which additional components contribute negligibly to the model. It is particularly useful when the variance or covariance decay is gradual, making the elbow point ambiguous.

In PCR, The Needle Algorithm identifies \( k \) by examining the second derivative of the cumulative explained variance function. Mathematically, the proportion of variance explained by the first \( k \) components is:
\begin{equation}
\text{Explained Variance}(k) = \frac{\sum_{i=1}^{k} \lambda_i}{\sum_{i=1}^{p} \lambda_i}
\end{equation}
where \( \lambda_i \) are the eigenvalues of \( X^T X \).

@@

For PLS, the Needle Algorithm follows the same principle but is applied to the explained covariance between \( X \) and \( Y \) instead of variance in \( X \). The proportion of covariance explained by the first \( k \) components is given by:
\begin{equation}
\text{Explained Covariance}(k) = \frac{\sum_{i=1}^{k} \text{Cov}(t_i, y)^2}{\sum_{i=1}^{p} \text{Cov}(t_i, y)^2}
\end{equation}
where \( t_i \) are the latent variables extracted by PLS.

@@

The Needle Algorithm provides a more formal alternative to the Elbow Method, particularly in cases where variance or covariance decays smoothly without a clear "elbow point".

\paragraph {Cross-Validation} \ \

Cross-Validation (CV) is a data-driven approach to selecting the optimal number of components. Unlike the Elbow Method or the Needle Algorithm, which focus on variance or covariance retention, CV directly evaluates the predictive performance of the model by minimizing the generalization error. This ensures that the selected number of components leads to the best tradeoff between bias and variance.

In PCR, the key challenge is to determine how many principal components (\( k \)) should be retained to ensure strong predictive performance. Since PCR does not use the response variable \( Y \) when selecting components, cross-validation is crucial in identifying the point where including additional components no longer improves prediction accuracy.

@@

Unlike PCR, PLS constructs components that maximize the covariance between \( X \) and \( Y \). Thus, CV in PLS plays a slightly different role: it helps determine the number of latent variables that maximize predictive accuracy.

@@

Since PLS incorporates \( Y \) in selecting components, it is expected to achieve lower optimal \( k^* \) and lower CV-MSE compared to PCR for a given dataset, particularly when \( X \) and \( Y \) are highly correlated.

While cross-validation is the most rigorous method for selecting \( k \), it has some practical considerations:
\begin{itemize}
    \item \textbf{Computational Cost}: Cross-validation is computationally expensive since it requires training multiple models for different values of \( k \) across multiple folds.
    \item \textbf{Bias-Variance Tradeoff}: Smaller \( k \) values lead to higher bias, while larger \( k \) values lead to higher variance. CV helps find the optimal balance.
    \item \textbf{Choice of \( K \)}: Typically, \( K = 5 \) or \( K = 10 \) is used. A smaller \( K \) reduces computation time, but larger \( K \) provides a more reliable error estimate.
\end{itemize}

The choice of \( k \) should be guided by a combination of these methods, balancing the need for variance retention and predictive accuracy. In empirical applications, cross-validation is typically preferred due to its ability to optimize for prediction error directly, therefore CV is applied in our simulations and empirical estimation.

@@

\paragraph{Severe Multicollinearity (Low Observations)} \ \

\begin{itemize}

    \item The dataset contains only \( n = 40 \) observations and \( p = 10 \) predictors, creating a scenario where the number of observations is significantly lower than in previous setups.
    \item A base predictor matrix is generated as:
    \[ X_{\text{gen}} \sim \mathcal{N}(0, I) \]
    \item Additional predictors are formed via a linear transformation:
    \[ X_{\text{additional}} = X_{\text{gen}} W + \mathcal{N}(0, 0.1) \]
    where \( W \) represents a randomly generated weight matrix, introducing strong collinearity among predictors.
    \item The response variable follows:
    \[ Y = X \beta + \epsilon, \quad \epsilon \sim \mathcal{N}(0, 1.0) \]
    \item Importantly, only \textbf{three} of the 10 predictors actively contribute to generating \( Y \), while the remaining seven are purely collinear noise variables.

\end{itemize}
